Matching Bullets

### 2015-11-23 imaging-paper.Rnw
working through pictures; exchanged subfig with subcaption

\usepackage{subfig}

@@

\begin{figure}[hbtp]
  \centering
    \subfloat[View from the side.\label{fig:sidex3p}]{%
      \includegraphics[width=\linewidth]{images/sidex3p.png}
    }
    \hfill
    \subfloat[View from the top.\label{fig:topx3p}]{%
    \includegraphics[width=\linewidth]{images/topx3p.png}
    }
\caption{An example of a scan of a bullet land.  \hh{XXX is the code for these pictures somewhere in the code folder? Let's put a link/note to it here. The aspect ratio of these 3D files should be adjusted such that it faithfully shows the actual geography. That should happen in the function plot3D.x3p.file }}
\end{figure}
Each fired bullet is provided in form of a set of six x3p files, where each file is a surface scan between adjacent grooves on the bullet, called a ``land". An example of plotting one of these lands is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}, which show side and top profiles of the land respectively.
\hh{The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artefact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clock-wise or anti-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic. XXX rotation in Ruger? twist rate?}

@@

<<fixedX, dependson='data', echo=FALSE, warning=FALSE, message=FALSE, fig.height=2.5, fig.width=6, out.width='\\linewidth', fig.pos='t!', fig.cap='Side profile of the height values of a bullet land across a fixed x value. Note that the global features dominate any deviations, which would correspond to markings.'>>=
library(RColorBrewer)
library(ggplot2)
library(scales)
library(dplyr)
library(x3pr)
library(x3prplus)

@@

qplot(data=subset(dbr111, x <= 100*1.5625^2 & x >= 99*1.5625^2), y, value, geom="line", size=I(1)) +
  geom_line(aes(x=xpred, y=ypred, group=x), 
            colour="grey30", size=0.25) +
  ylab(expression(paste("Surface Measurements (in ",mu,m,")", sep=""))) + 
  theme_bw() + 
  theme(legend.position="bottom") + coord_equal()
@

@@

One reasonable approach is to fit a circle to the curve and compute the residuals from this fit. The scatterplot below shows the residuals of such a fit. The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For cross sections of $x$ values between 75 and 80, there is a residual circular structure that does not show up for all cross sections.

<<residual, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\linewidth'>>=
qplot(data=subset(dbr111, x <= 80*1.5625^2 & x >=75*1.5625^2), y, value-ypred, colour=factor(x), geom="line", size=I(1)) +
  scale_colour_brewer("x", palette="Paired") + 
  theme_bw() + 
  theme(legend.position="bottom")
@

<<residual2, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\linewidth'>>=
qplot(data=subset(dbr111, x <= 5*1.5625^2), y, value-ypred, colour=factor(x),
      geom="line", size=I(1)) +
  scale_colour_brewer("x", palette="Paired") + 
  theme_bw() + 
  theme(legend.position="bottom")
@

@@

lines <- matches %>% group_by(lineid) %>% summarise(
  meany = mean(y, na.rm=T),
  miny = min(y, na.rm=T),
  maxy = max(y, na.rm=T) + 1.5625,
  match = isMatch(lineid, type),
  type = type1[1]
)
lines <- subset(lines, lineid != 0)
@
\begin{figure}
\centering
    \subfloat[Crosscuts at $x = 100$.\label{fig:crosscut}]{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='.49\\textwidth'>>=
qplot(y, resid, group=bullet, colour=bullet, data=lof, 
      geom="line", alpha=I(0.6)) + 
  scale_colour_brewer("", palette="Set1") + theme_bw() +
  theme(legend.position = c(.9, .2))
@
    }
    \hfill
    \subfloat[Adjusted (Bullet 1-5 horizontally shifted) crosscuts at $x = 100$.\label{fig:crosscutX}]{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='.49\\textwidth'>>=
# qplot(y, resid, group=bullet, colour=bullet, data=lofX, 
#       geom="line", alpha=I(0.6)) + 
#   scale_colour_brewer(palette="Set1") + theme_bw() +
#   theme(legend.position="none")
qplot(y, resid, group=bullet, colour=bullet, data=lofY, 
      geom="line", alpha=I(0.6)) + 
  scale_colour_brewer(palette="Set1") + theme_bw() +
  theme(legend.position="none")
@
    }
\caption{\label{fig:cross100} Lines showing surface crosscuts at $x = 100$. A horizontal shift of the values of Bullet 1-5 to the right shows the similarity of the striation marks.}
\end{figure}

@@

\begin{figure}
\centering
    \subfloat[Loess smooth at $x = 100$ (span is 0.03).\label{fig:smooth}]{%
<<smooth, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3.5, out.width='.49\\textwidth', warning=FALSE>>=
qplot(data=lofX, y, resid, colour=bullet, size=I(0.75), geom="line", group=bullet) +
  geom_line(aes(y=l30), colour="grey30", size=0.5) +
  facet_grid(bullet~.) +
  scale_colour_brewer("", palette="Set1") +
  theme_bw() +
  theme(legend.position="none")
@
    }
    \hfill
    \subfloat[Values above/below a threshold level are defined as peaks and grooves. \label{fig:smoothcutb}]{%
<<smoothcut, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=4, out.width='.49\\textwidth', warning=FALSE>>=
qplot(data=lofX, x=y, y=l30, colour=type, geom="line", 
      group=bullet, size = I(0.75)) + 
  facet_grid(bullet~.) +
  ylab("Individual characteristics") + 
#  geom_point(aes(y=r05)) +
  geom_rect(xmin = 200, xmax = 2260, ymin = -.75, ymax = 0.75, colour=NA, fill=alpha("grey90", alpha =0.5)) +
  scale_colour_brewer(palette="Set2", na.value = NA) +
  theme_bw() +
  theme(legend.position="bottom")
@
    }
\hfill
    \subfloat[Grey backgrounds identify a line on one of the bullets. Overlaps of the same color indicate matching striation marks.
    Matching striations are marked by an `o', mismatches are marked by an `x'.   \label{fig:smoothcutd}]{%
<<smoothmatch, echo=FALSE, dependson='twolands100match', fig.width=7, fig.height=2.75, out.width='.49\\textwidth'>>=
ggplot() +
  geom_rect(aes(xmin = miny, xmax = maxy), ymin = 0.25, ymax=2.75, fill="grey80", data = subset(lines, lineid != 0)) +
  geom_tile(data=lofX, aes(x=y, y=bullet, fill=type)) + 
  scale_y_discrete(expand = c(0.2,0)) +
  scale_fill_brewer(palette="Set2") + 
  theme_bw() +
  ylab("") +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= .4, label= "x", data = subset(lines, !match & lineid !=0)) +
  geom_text(aes(x = meany), y= .4, label= "o", data = subset(lines, match & lineid !=0))
@
    }    
\caption{\label{fig:match} Matching striation marks: smooth (a), cut (b), and match (c).}
\end{figure}

@@

\section{Circular Fit}

Assume, $n$ data points $(x_1, y_1), (x_2, y_2), ..., (x_n, y_n)$ are given that are (approximately) located on a circle. We want to estimate the location of the center and radius of the best fitting circle.

\noindent
The  derivate of $D$ with respect to $r$ is:
\begin{eqnarray*}
\frac{d}{dr} D &=& 2 \sum_i \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right) 2 r = \\
&=& 4 r \left( n r^2 - \sum_i (x_i-a)^2 - \sum_i(y_i-b)^2 \right) 
\end{eqnarray*}
For a minimum then holds:
\begin{equation}\label{eq:rmin}
\frac{d}{dr} D = 0 \stackrel{r \neq 0}{\iff} nr^2  = \sum_i (x_i-a)^2 + \sum_i(y_i-b)^2 
\end{equation}
%
%
The  derivative of $D$ with respect to $a$ is:
\begin{eqnarray*}
\frac{d}{da} D &=& 2 \sum_i \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right) 2 (x_i - a) = \\
&=& -4 \left[ a \cdot nr^2 + \sum_i (x_i - a)^3  + \sum_i (x_i - a) (y_i - b)^2 \right]
\end{eqnarray*}
Using (\ref{eq:rmin}) for $nr^2$  in the equation above we get:
\begin{eqnarray*}
\frac{d}{da} D &=& -4 \left[  \sum_i a(x_i-a)^2 +  \sum_i a(y_i-b)^2  + \right. \\
&& \phantom{-4 \ \ } \left . \sum_i (x_i - a)^3  + \sum_i (x_i - a) (y_i - b)^2 \right]  = \\
&=& -4 \left[ \sum_i (x_i-a)^2 (a + x_i - a)  + \sum_i (x_i - a + a) (y_i - b)^2 \right] = \\
&=& -4 \left[ \sum_i (x_i-a)^2 x_i   + \sum_i x_i  (y_i - b)^2 \right] \stackrel{\sum_i x_i = \sum_i y_i = 0}{=} \\
&=& -4 \left[ \sum_i x_i^3   + \sum_i x_i y_i^2  - 2a s_{xx} - 2b s_{xy} \right], 
\end{eqnarray*}
where $s_{xx} = \sum_i x_i^2, s_{xy} = \sum_i x_i y_i$ and $s_{yy} = \sum_i y_i^2$.

\noindent
Likewise, we get for the derivative of $D$ with respect to $b$:
\begin{eqnarray*}
\frac{d}{db} D &=& -4 \left[ \sum_i y_i^3   + \sum_i x_i^2 y_i - 2a s_{xy} - 2b s_{yy} \right] 
\end{eqnarray*}
For the minimum we therefore get a system of two linear equations in $a$ and $b$:
\begin{eqnarray*}
2 s_{xx} a + 2 s_{xy} b = c_1 && \text{ with } c_1 = \sum_i x_i^3 + x_i y_i^2 \\
2 s_{xy} a + 2 s_{yy} b = c_2 &&\text{ with } c_2 = \sum_i x_i^2 y_i + y_i^3.
\end{eqnarray*}
This resolves to 
\begin{eqnarray*}
\hat{a} &=& \frac{c_1 s_{yy} - c_2 s_{xy}}{2 s_{xx} s_{yy} - 2 s_{xy}^2},\\
\hat{b} &=& \frac{c_2 s_{xx} - c_1 s_{xy}}{2 s_{xx} s_{yy} - 2 s_{xy}^2}, \text{ and}\\
\hat{r^2} &=& \frac{1}{n}s_{xx} + \frac{1}{n}s_{yy} + \hat{a}^2 + \hat{b}^2.
\end{eqnarray*}
\end{document}

>>>>

\usepackage{subcaption}

@@

\begin{figure}[hbtp]
  \centering
\begin{subfigure}[b]{\linewidth}
\caption{View from the side.\label{fig:sidex3p}}{%
      \includegraphics[width=\linewidth]{images/sidex3p.png}
    }
\end{subfigure}    
\begin{subfigure}[b]{\linewidth}
    \caption{View from the top.\label{fig:topx3p}}{%
    \includegraphics[width=\linewidth]{images/topx3p.png}
    }
\end{subfigure}
\caption{An example of a scan of a bullet land from one groove to the next. }
\end{figure}
Each fired bullet is provided in form of a set of six x3p files, where each file is a surface scan between adjacent grooves on the bullet, called a ``land". An example of plotting one of these lands is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}, which show side and top profiles of the land respectively.
\hh{The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artefact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clock-wise or anti-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic. XXX rotation in Ruger? twist rate?}

@@

\begin{figure}
<<fixedX, dependson='data', echo=FALSE, warning=FALSE, message=FALSE, fig.height=2, fig.width=6, out.width='\\linewidth'>>=
library(RColorBrewer)
library(ggplot2)
library(scales)
library(dplyr)
library(x3pr)
library(x3prplus)

@@

qplot(data=subset(dbr111, x <= 100*1.5625^2 & x >= 99*1.5625^2), y, value, geom="line", size=I(1)) +
  geom_line(aes(x=xpred, y=ypred, group=x), 
            colour="grey30", size=0.25) +
#  ylab(expression(paste("Surface Measurements (in ",mu,m,")", sep=""))) + 
  ylab("") +
  theme_bw() + 
  theme(legend.position="bottom") #+ coord_equal()
@
\caption{\label{fig:fixedX}Side profile of the surface measurements (in $\mu m$) of a bullet land across a fixed $x$ value. Note that the global features dominate any deviations, corresponding to the individual characteristics of striation marks.}
\end{figure}

@@

One reasonable approach is to fit a circle to the curve and compute the residuals from this fit.

For this, assume that $n$ data points $(x_1, y_1), (x_2, y_2), ..., (x_n, y_n)$ are given that are (approximately) located on a circle. We want to estimate the location of the center and radius of the best fitting circle using a least squares approach.

\noindent
The  derivate of $D$ with respect to $r$ is:
\begin{eqnarray*}
\frac{d}{dr} D &=& 2 \sum_i \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right) 2 r = \\
&=& 4 r \left( n r^2 - \sum_i (x_i-a)^2 - \sum_i(y_i-b)^2 \right) 
\end{eqnarray*}
For a minimum then holds:
\begin{equation}\label{eq:rmin}
\frac{d}{dr} D = 0 \stackrel{r \neq 0}{\iff} nr^2  = \sum_i (x_i-a)^2 + \sum_i(y_i-b)^2 
\end{equation}
%
%
The  derivative of $D$ with respect to $a$ is:
\begin{eqnarray*}
\frac{d}{da} D &=& 2 \sum_i \left( r^2 - (x_i-a)^2 - (y_i-b)^2 \right) 2 (x_i - a) = \\
&=& -4 \left[ a \cdot nr^2 + \sum_i (x_i - a)^3  + \sum_i (x_i - a) (y_i - b)^2 \right]
\end{eqnarray*}
Using (\ref{eq:rmin}) for $nr^2$  in the equation above we get:
\begin{eqnarray*}
\frac{d}{da} D &=& -4 \left[  \sum_i a(x_i-a)^2 +  \sum_i a(y_i-b)^2  + \right. \\
&& \phantom{-4 \ \ } \left . \sum_i (x_i - a)^3  + \sum_i (x_i - a) (y_i - b)^2 \right]  = \\
&=& -4 \left[ \sum_i (x_i-a)^2 (a + x_i - a)  + \right.\\
&& \phantom{-4 \ \ } \left .\sum_i (x_i - a + a) (y_i - b)^2 \right] = \\
&=& -4 \left[ \sum_i (x_i-a)^2 x_i   + \sum_i x_i  (y_i - b)^2 \right] 
\stackrel{\begin{array}{c}\sum_i x_i = 0\\
\sum_i y_i = 0\end{array}}{=} \\
&=& -4 \left[ \sum_i x_i^3   + \sum_i x_i y_i^2  - 2a s_{xx} - 2b s_{xy} \right], 
\end{eqnarray*}
where $s_{xx} = \sum_i x_i^2, s_{xy} = \sum_i x_i y_i$ and $s_{yy} = \sum_i y_i^2$.

\noindent
Likewise, we get for the derivative of $D$ with respect to $b$:
\begin{eqnarray*}
\frac{d}{db} D &=& -4 \left[ \sum_i y_i^3   + \sum_i x_i^2 y_i - 2a s_{xy} - 2b s_{yy} \right] 
\end{eqnarray*}
For the minimum we therefore get a system of two linear equations in $a$ and $b$:
\begin{eqnarray*}
2 s_{xx} a + 2 s_{xy} b = c_1 && \text{ with } c_1 = \sum_i x_i^3 + x_i y_i^2 \\
2 s_{xy} a + 2 s_{yy} b = c_2 &&\text{ with } c_2 = \sum_i x_i^2 y_i + y_i^3.
\end{eqnarray*}
This resolves to 
\begin{eqnarray*}
\hat{a} &=& \frac{c_1 s_{yy} - c_2 s_{xy}}{2 s_{xx} s_{yy} - 2 s_{xy}^2},\\
\hat{b} &=& \frac{c_2 s_{xx} - c_1 s_{xy}}{2 s_{xx} s_{yy} - 2 s_{xy}^2}, \text{ and}\\
\hat{r^2} &=& \frac{1}{n}s_{xx} + \frac{1}{n}s_{yy} + \hat{a}^2 + \hat{b}^2.
\end{eqnarray*}

The scatterplot in Figure~\ref{fig:residual} shows the residuals of such a fit. The residuals are dominated, as expected, by the grooves, which show up as large positive residuals. For a surface cross cut at $x = 100$ there is a residual circular structure that does not show up for all cross sections.

\begin{figure}
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:residuala}Residual structure of a surface crosscut at $x = 1.5625$ (bottom of the bullet).}
<<residual2, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\linewidth'>>=
qplot(data=subset(dbr111, x <= 1.5625), y, value-ypred, #colour=factor(x),
      geom="line", size=I(1)) +
#  scale_colour_brewer("x", palette="Paired") + 
  theme_bw() + 
  theme(legend.position="bottom")
@
\end{subfigure}    
\begin{subfigure}[t]{\linewidth}
\caption{\label{fig:residualb} Residual structure of a surface crosscut at $x = 100.00$}
<<residual, dependson='fixedX', echo=FALSE, warning=FALSE, fig.height=3, out.width='\\linewidth'>>=
#qplot(data=subset(dbr111, x <= 80*1.5625^2 & x >=75*1.5625^2), y, value-ypred,
qplot(data=subset(dbr111, x == 100), y, value-ypred,  #colour=factor(x), 
      geom="line", size=I(1)) +
  scale_colour_brewer("x", palette="Paired") + 
  theme_bw() + 
  theme(legend.position="bottom")
@
\end{subfigure}
\caption{\label{fig:residual} Residual structure of  circular fits at two different cross-section. The residual structure in (b) shows systematic structure, indicating that the fit is not entirely appropriate. }
\end{figure}

@@

lines <- matches %>% group_by(lineid) %>% summarise(
  meany = mean(y, na.rm=T),
  miny = min(y, na.rm=T),
  maxy = max(y, na.rm=T) + 1.5625,
  match = isMatch(lineid, type),
  type = type1[1]
)
lines <- subset(lines, lineid != 0)
@
\begin{figure}
\centering
\begin{subfigure}[t]{\linewidth}
\caption{Crosscuts at $x = 100$.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='.49\\textwidth'>>=
qplot(y, resid, group=bullet, colour=bullet, data=lof, 
      geom="line", alpha=I(0.6)) + 
  scale_colour_brewer("", palette="Set1") + theme_bw() +
  theme(legend.position = c(.9, .2))
@
    }
\end{subfigure}    
\begin{subfigure}[t]{\linewidth}
    \caption{Adjusted (Bullet 1-5 horizontally shifted) crosscuts at $x = 100$.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='.49\\textwidth'>>=
# qplot(y, resid, group=bullet, colour=bullet, data=lofX, 
#       geom="line", alpha=I(0.6)) + 
#   scale_colour_brewer(palette="Set1") + theme_bw() +
#   theme(legend.position="none")
qplot(y, resid, group=bullet, colour=bullet, data=lofY, 
      geom="line", alpha=I(0.6)) + 
  scale_colour_brewer(palette="Set1") + theme_bw() +
  theme(legend.position="none")
@
    }
\end{subfigure}
\caption{\label{fig:cross100} Lines showing surface crosscuts at $x = 100$. A horizontal shift of the values of Bullet 1-5 to the right shows the similarity of the striation marks.}
\end{figure}

@@

\begin{figure}
\centering
    \begin{subfigure}[t]{\linewidth}
\caption{Loess smooth at $x = 100$ (span is 0.03).\label{fig:smooth}}{%
<<smooth, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3.5, out.width='.49\\textwidth', warning=FALSE>>=
qplot(data=lofX, y, resid, colour=bullet, size=I(0.75), geom="line", group=bullet) +
  geom_line(aes(y=l30), colour="grey30", size=0.5) +
  facet_grid(bullet~.) +
  scale_colour_brewer("", palette="Set1") +
  theme_bw() +
  theme(legend.position="none")
@
    }
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{Values above/below a threshold level are defined as peaks and grooves. \label{fig:smoothcutb}}{%
<<smoothcut, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=4, out.width='.49\\textwidth', warning=FALSE>>=
qplot(data=lofX, x=y, y=l30, colour=type, geom="line", 
      group=bullet, size = I(0.75)) + 
  facet_grid(bullet~.) +
  ylab("Individual characteristics") + 
#  geom_point(aes(y=r05)) +
  geom_rect(xmin = 200, xmax = 2260, ymin = -.75, ymax = 0.75, colour=NA, fill=alpha("grey90", alpha =0.5)) +
  scale_colour_brewer(palette="Set2", na.value = NA) +
  theme_bw() +
  theme(legend.position="bottom")
@
    }
\end{subfigure}    
\begin{subfigure}[t]{\linewidth}
    \caption{Grey backgrounds identify a line on one of the bullets. Overlaps of the same color indicate matching striation marks.
    Matching striations are marked by an `o', mismatches are marked by an `x'.   \label{fig:smoothcutd}}{%
<<smoothmatch, echo=FALSE, dependson='twolands100match', fig.width=7, fig.height=2.75, out.width='.49\\textwidth'>>=
ggplot() +
  geom_rect(aes(xmin = miny, xmax = maxy), ymin = 0.25, ymax=2.75, fill="grey80", data = subset(lines, lineid != 0)) +
  geom_tile(data=lofX, aes(x=y, y=bullet, fill=type)) + 
  scale_y_discrete(expand = c(0.2,0)) +
  scale_fill_brewer(palette="Set2") + 
  theme_bw() +
  ylab("") +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= .4, label= "x", data = subset(lines, !match & lineid !=0)) +
  geom_text(aes(x = meany), y= .4, label= "o", data = subset(lines, match & lineid !=0))
@
    }
\end{subfigure}
\caption{\label{fig:match} Matching striation marks: smooth (a), cut (b), and match (c).}
\end{figure}

@@

\end{document}
